---
title: "Riri in the Wild Zone"
subtitle: "A Distant Reading of Sapphic Urban Latin Music"
author: "Laura Isabel Sastoque Pabón"

format:
  pdf:
    toc: true
    embed-resources: true
    code-fold: true
    link-external-newwindow: true
    mainfont: Garamond
    
execute:
  warning: false

from: markdown+emoji 
---

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: load-pckgs-data

# load required packages

library(tidyverse)
library(tidytext)
library(readtext)
library(readxl)
library(syuzhet)
library(viridis)
library(gt)
library(quanteda)
library(quanteda.textmodels)
library(quanteda.textplots)
library(quanteda.textstats)
library(patchwork)

# set seed
set.seed(20022901)

# load required objects
load("corpus_objects/metadata.rda")
metadata <- corpus_metadata
rm(corpus_metadata)

```

::: callout-important
Follow this link for a full GitHub Repository of this project

Final Project Repository
:::

## 1 **Introduction**

In recent years, distant reading methods have earned popularity for the analysis of literary texts. These methods hinge on the transformation of text into features— data that can be quantified, qualified, and visualized using computational tools.[^1] Often framed in methodological opposition to close reading, distant reading has been scrutinized for its impulse to automate interpretation, and as its name indicates, to build distance between the reader and the text. However, scholars such as Katherine Bode have theorized that distant reading does not stand in opposition to close reading, but is rather a conceptual legacy of the latter, which prioritizes the text as the ultimate source of meaning.[^2] Furthermore, scholars have called attention to distant reading methods for their ability to reduce the reader’s bias when working with texts, and for their potential to expand existing literary canons through a wider, more transparent inquiry of literary material.[^3] So far, the majority of applications of distant reading in the digital humanities have had literary texts as their objects of inquiry—novels, poems, even television scripts.[^4] However, what happens when the objects of a distant reading study are not commonly considered “literary”? What happens when these texts are, in fact, not even meant to be read, but instead heard, and danced to?

[^1]: Maciej Eder, et al. “Stylometry with R: A Package for Computational Text Analysis.” *The R Journal 8* (January 1, 2016): 107–21. https://doi.org/10.32614/RJ-2016-007.

[^2]: Katherine Bode. “The Equivalence of ‘Close’ and ‘Distant’ Reading; or, Toward a New Object for Data-Rich Literary History.” *Modern Language Quarterly* (Seattle) 78, no. 1 (2017): 77–106. https://doi.org/10.1215/00267929-3699787.

[^3]: Natalie M. Houston. “Distant Reading and Victorian Women’s Poetry.” In *The Cambridge Companion to Victorian Women’s Poetry,* 249–65. Cambridge University Press, 2019. https://doi.org/10.1017/9781316856543.017.

[^4]: Joanna Byszuk. “The Voices of Doctor Who – How Stylometry Can Be Useful in Revealing New Information About TV Series.” *Digital Humanities Quarterly* 014, no. 4 (December 20, 2020).

In fact, few studies have ventured into the realm of music wearing a distant reading toolbelt. Those that do, contain little critical reflection on the application of these methods to this new textual manifestation.[^5] This essay aims to embark on the experimental task of conducting a distant reading of sapphic urban Latin music with an open, yet critical lens. In recent years, urban Latin music—known for its highly sexualized lyrics and male-dominated industry—has been undergoing a transformation.[^6] Not only are female artists earning more and more notoriety, but a subset of queer, and particularly sapphic artists is also emerging.[^7] In a genre where the hyper-sexualization of women is a convention, it is pertinent to interrogate whether women singers are appropriating or modifying the language. Do sapphic songs— particularly, at a lyrical level— frame desire in a similar way to straight male artists, or are they creating a new lexical tradition that breaks pre-existing conventions?

[^5]: See for instance Elizabeth Monzingo and Daniel Shanahan. “The Expression of Self and Grief in the Nineteenth Century: An Analysis through Distant Readings.” *Nineteenth-Century Music Review* 18, no. 1 (2021): 83–107. https://doi.org/10.1017/S1479409819000697.

[^6]: Henceforth, urban Latin music is used as an umbrella term for reggaetón, trap Latino, dembow, bachata, merengue, among other styles used by the artists in the corpus.

[^7]: In this essay, the use of the term “sapphic” is an intentional choice to encompass the variety of identities through which queer desire aimed at women presents itself in urban Latin music. This term is preferred as opposed to “lesbian,” since it includes artists who do not exclusively sing about women but have a few select songs that were included in the corpus. These artists include Cazzu, Maria Becerra, Karol G, among others. It is also preferred to the word “queer,” since this essay does not encompass male artists who write about other men, who are also an emerging subset in the genre, albeit less recognized by the mainstream. The word “sapphic” is also meant to include transgender women who at times sing about other women, such as Villano Antillano.

By conducting distant reading analyses such as text mining, this study aims to understand the appropriations and transformations of language that sapphic singers are utilizing to advance their specific agenda within the genre. This study conducts this analysis, however, with the knowledge that the methods of distant reading were not built to deal with these types of texts.[^8] Urban Latin music song lyrics are slang-heavy, filled with non-standard manipulations of the Spanish language. They are repetitive, highly onomatopoeic, and at times play with the boundaries between Spanish, English, and Portuguese.[^9] Therefore, this essay will not take distant reading as a direct path to knowledge-making but will instead engage in a two-way negotiation with its affordances, shortcomings, and future possibilities. In doing so, it argues that sapphic artists are able to simultaneously imitate and eschew the male-dominated language of desire in urban Latin music. In Elaine Showalter’s terms, the particular position of sapphic urban Latin music as an emerging, yet highly genealogical subgroup, allows it to stand in a superposition of the “muted zone” and the “wild zone” of women’s writing, distinguished by their relationality to male convention.

[^8]: Distant reading practices are, first of all, most advanced for the English language. Secondly, they are based on dictionaries compiled, in their majority, from newspaper sources, which contain fairly standard manifestations of language. See Hans Ole Hatzel, et al. “Machine Learning in Computational Literary Studies.” *It - Information Technology* 65, no. 4–5 (August 1, 2023): 200–217. <https://doi.org/10.1515/itit-2023-0041>, pp. 201-202.

[^9]: Patricio Josué Velarde Dediós. “Reguetón de mujeres para mujeres: dos caminos de empoderamiento.” *Antec: Revista Peruana de Investigación Musical* 5, no. 1 (August 14, 2021): 91–109. https://doi.org/10.62230/antec.v5i1.112.

## 2 **Sapphic Desire in Urban Latin Music**

In the last decade, women artists have slowly been carving out a space for themselves in urban Latin music. Previous literature has meditated on the role of some of the pioneering women *reggaetoneras*, such as Ivy Queen.[^10] In a genre that blatantly advances men’s utilization of women as mediums for interrelationality, and as sexual, often less-than-human objects, women artists have had to carefully tip-toe the conventions of the genre to cultivate their careers. In their earliest interventions, women were relegated to uncredited lyrical accompaniments or ad-libs in male-dominated songs. In some of their early musical projects, women artists would often accommodate to convention and portray themselves as the objects of male desire.[^11] This was perhaps with the exception of Ivy Queen, whose songs were marked by a reclamation of her own sexual agency. A “queer” gender presentation, accompanied with a low-pitched voice, also characterized her strategy.[^12] As Patricio Josué Velarde Dediós points out, this trend has shifted since the mid-2010s, with artists such as Becky G, Natti Natasha, and Karol G. This new wave is characterized by high levels of collaboration amongst women artists, whereas before they would largely collaborate with men. Furthermore, musical projects directly challenge some of the dualities and encasings that women face in reggaetón. Such is the case of Becky G’s *Mala Santa* album, which aims to challenge the virgin/whore dichotomy.

[^10]: Dara E. Goldman. “Walk like a Woman, Talk like a Man: Ivy Queen’s Troubling of Gender.” *Latino Studies* 15, no. 4 (November 1, 2017): 439–57. https://doi.org/10.1057/s41276-017-0088-5.

[^11]: Velarde Dediós, p. 92.

[^12]: Goldman, p. 448.

Most recently, a further development in this trend of women’s participation has begun—the inclusion of sapphic artists. This has not only signified the popularization of new, lesbian-identifying artists such as Young Miko, but also the release of sapphic songs by artists who mostly sing about heterosexual desire, such as with Karol G and Kali Uchis’ “Labios Mordidos.” This development poses new theoretical challenges. How can women establish themselves as both subjects and objects of desire in a sphere that was built through the homosocial relations of men? To an extent, the answer to this question is experimentation. Reggaetón bears very particular linguistic characteristics—from the hegemonic character of Puerto Rican slang to an established set of zoomorphic words to refer to women.[^13] Part of the way sapphic artists are inserting themselves in this genre is by bending these conventions or imagining new possibilities for the lexical space of reggaetón.

[^13]: Pavlina Vankova. “Studying the Vocabulary of Reggaeton Song Lyrics.” *Topics in Linguistics* 23, no. 2 (2022): 63–88. <https://doi.org/10.2478/topling-2022-0012>; Angel Torres-Toukoumidis, et al. “Computational Analysis of Latin Music Songs Through Tokenization. Case of Female Artists and Reggaeton.” In *Communication and Applied Technologies*, edited by Paulo Carlos López-López, et al, 527–35. Smart Innovation, Systems and Technologies. Singapore: Springer Nature, 2023. https://doi.org/10.1007/978-981-19-6347-6_47.

It is this very space of inventiveness that theorists of sapphic desire have previously encountered. In “Refiguring Lesbian Desire,” Elizabeth Grosz engages in an experimental reflection of how we can reframe lesbian desire away from male-centric, often phallocentric conceptions of “lack” and “absence”—of a gaping void waiting to be filled. Grosz associates this way of thinking with Freud’s psychoanalytic theories, which conceive of desire in negative terms. Instead, Grosz proposes a theory of desire borrowing from rhizomatic thinking, where it is defined in terms of “parts and bits and interconnections, \[what\] is new, \[what\] is exploratory, that opens up further spaces, induces further intensities, speeds up, enervates, and proliferates production (production of the body, production of the world).”[^14] In these terms, desire becomes a productive force, a force that does not incur hierarchy but rather flattens it. A force that inevitably engenders knowledge.[^15]

[^14]: Elizabeth Grosz. “Refiguring Lesbian Desire.” In Space, Time and Perversion. Routledge, 1995, p. 81.

[^15]: Grosz hints at this conclusion in the ominous ending to her essay, “that is the most we can hope for from knowledge. Or desire” (p. 81).

This productive space of sapphic desire is perhaps what Elaine Showalter was imagining when she theorized the “wild zone” of women’s culture. This is the space in women’s experience, in their way of understanding and engaging with the world, that is “no man’s land.” Because all spheres of society are mediated by male conventions, including language, the wild zone might even be a space of foreign tongues, a space outside the “prison-house” of language.[^16] There is no corresponding wild zone for men, since their conventions and structures already inform language as we know it, and thus bear no mystery.[^17] According to Showalter, the wild zone is just a metaphorical proposition, because in reality most of women’s writing—their production—is mediated through the “muted zone.” In the muted zone, “muted groups must mediate their beliefs through the allowable forms of dominant structures.”[^18] Hence, Showalter proposes that we think of women’s writing as a double-voiced discourse, that toggles the line between the muted and wild zones, and that goes beyond the dualistic view that women writers either imitate their predecessors or revise them.[^19]

[^16]: Elaine Showalter. “Feminist Criticism in the Wilderness.” *Critical Inquiry* 8, no. 2 (1981): 179–205, p. 194.

[^17]: It must acknowledged that this theory represents gender in strictly binary terms, which does not align with the gender identifications of many of the artists included in the corpus. However, this theory is useful in its theorizing of a space of language—and naturally, desire—where men are not allowed access.

[^18]: Showalter, p. 201.

[^19]: Showalter, p. 204.

When writing in the sphere of urban Latin music, women artists are confronted with existing scripts of desire. In the “prison-house” of urban Latin lyrics, to express desire for a woman almost inevitably signifies her sexualization and reduction to non-human elements. This conundrum is palpable in many of the recent hits of sapphic artists. In an interview for *Rolling Stone*, Young Miko discussed her debut EP *Trap Kitty* (2022). The songs follow Riri, a pole dancer working at a strip club in Puerto Rico. Upon first sight, one may associate the way Young Miko sings about Riri with the way most men in the genre sing about women. In the song “Riri,” Young Miko sings:

+---------------------------------------------------+-----------------------------------------------------------------------+
| *Le toco su cuerpo, eso es lo que ella ama,*\     | I touch her body, that's what she loves,\                             |
| *y prende la webcam cuando está en su cama.*\     | and she turns on her webcam when she's in bed\                        |
| *(...)*\                                          | (...)\                                                                |
| *Ella se va viral en IG cuando publica*\          | She goes viral on Instagram when she posts,\                          |
| *Me dijo que sus captions siempre me los dedica*\ | she told me her captions are all for me.\                             |
| *(...)*\                                          | (...)\                                                                |
| *Pa' ver su contenido tienes que pagar el prime.* | To watch her content, you must pay for a prime \[subscription\].[^20] |
+---------------------------------------------------+-----------------------------------------------------------------------+

[^20]: Translated by the author.

To a listener familiar with popular urban Latin songs, these lyrics are reminiscent of a common trend in male-dominated songs—the overt discussion of sexual work almost as a desirable trait in women.  One such song is “Only Fans” by Young Martino, Myke Towers, and Lunay.[^21]

[^21]: I would invite readers to watch the music video for this song, to get an even more apparent insight on the images of women advanced by these lyrics.

+--------------------------------------+------------------------------------------+
| *Con el tiempo vi un vídeo por ahí*\ | In time, I saw a video laying around,\   |
| *Y en OnlyFans rápido me suscribí*\  | and quickly subscribed to her OnlyFans.\ |
| *(...)*\                             | (...)\                                   |
| *Y e' una santa en Instagram (así)*\ | She's a saint on Instagram,\             |
| *Y e' una puta en OnlyFans (¡ah!)*   | but a whore on OnlyFans.[^22]            |
+--------------------------------------+------------------------------------------+

[^22]: Translated by the author.

A non-attentive reader would be quick to group “Riri” among the set of women’s songs that primarily imitate men’s language of desire. However, Young Miko asserts that her music— although “raunchy”— is meant for women and queer folks.[^23] And upon closer inspection, it is possible to glean some of this intentionality. Whereas “Only Fans” is largely about the male singer “I,” “Riri” predominantly endows its female character with agency, speaking of what she loves, and only once using a reference to the “I” (even then, the pronoun is in its reflexive form ‘me’ + verb). How many more of these subtle differences, these delicate negotiations between the muted and wild zones would we encounter if we were able to look at a sapphic urban Latin music as a whole? Or, in realistic terms, if we looked at a representative sample of the emerging phenomenon of sapphic music, and examined it against a set of conventional, male-dominated songs? This is where the methodologies of distant reading prove useful for the question at hand. By building, quantifying, and visualizing a corpus of song lyrics, we will engage with sapphic music’s *prise de la parole,* which goes hand-in-hand with women writers’ *prise de conscience.*[^24]

[^23]: Solá Santiago Frances “Welcome to Young Miko’s World.” *Rolling Stone* (blog), June 26, 2023. <https://www.rollingstone.com/music/music-features/young-miko-trap-kitty-bad-bunny-karol-g-1234771397/>.

[^24]: Showalter, p. 190.

## 3 **The Creation of a Corpus-Canon**

The selection of a representative corpus is a key step in any distant reading analysis. This entity—the “corpus”—stems from the field of corpus linguistics, which aims “to construct and empirically analyze a sample of texts as the basis for the meaningful generalizations about a language or discourse domain.”[^25] The corpus describes not the ultimate goal of inquiry, but its method or material. The corpus has a target—in this case, the domain of sapphic urban Latin music. An imperfect analogue in close-reading-based literary studies is perhaps the literary canon. Although a canon is constructed through a historical process of scholarly inquiry, rather than statistical sampling, it decidedly positions certain texts as “representative” of a movement or domain of literature. The literary canon carries a certain air of authority that attracts scholars seeking to understand a specific time period or moment in literary history. This is similar to the process by which a corpus often gets reused by authors seeking to reproduce or build on past scholars’ contributions. Natalie Houston builds on Annette Kolodny’s ideas in stating that computational feminist criticism allows for “explicit reflection on the decisions made in conducting the dataset.”[^26] For Houston, distant reading offers a departure away from literary canons that are often examined uncritically. Showalter reaches a similar point, in stating that feminist criticism at its core aimed to resist existing canons.[^27] Therefore, scholars of computational feminist criticism believe that a corpus, by making the boundaries of selection explicit, bears advantages of representativity over a literary canon.[^28]

[^25]: Jesse Egbert, et al. *Designing and Evaluating Language Corpora: A Practical Framework for Corpus Representativeness*. Cambridge University Press, 2022. <https://doi.org/10.1017/9781316584880>, pp. 5-6.

[^26]: Houston, p. 251.

[^27]: Showalter, p. 182.

[^28]: Houston, p. 255.

Nonetheless, the boundary between a corpus and a canon is one that should be threaded carefully. As theorists of corpus linguistics have suggested, it is perhaps just as easy to introduce bias into a corpus as it is to drown in the incompleteness of a canon.[^29] This realization was central in the process of creating a corpus for this study. Delineating a corpus of sapphic songs in urban Latin music was a challenging task. The artists are multifaceted—they differ in terms of age, nationality, sexual identity, and artistic goals. Likewise, the peculiarities of their lyrics sit in discomfort with the requirements of a corpus analysis— their texts are highly oral, repetitive, and slang-heavy. In other words, even though sapphic urban Latin music is bound together by a shared identity, it seems to resist corpus-making, and by extension, canon-making. Part of the self-reflective process of dataset creation was an active evaluation of personal biases in artist-song selection, and an active effort to counteract or balance these impulses. Appropriate selection criteria for sapphic urban Latin songs are more complicated than looking at a chart of the most popular artists and scraping their top songs. A selection based on popularity would erase the work of smaller, emerging artists, with rich and varying agendas and lexical possibilities. To select artists based on popularity would be to conform to a canon advanced by mainstream music industries, the same canon that women’s writing has for centuries attempted to resist. Therefore, the corpus-creation process for this study was a constructive one. It began with the charts and thus selected top songs by popular artists such as Young Miko, Karol G, and María Becerra. Then, it took the time to scrutinize other areas where emerging sapphic artists often promote their music. This included social media and Spotify playlists and led to the inclusion of smaller artists such as Chzter and Gudnana, as well as artists whose agenda is slightly more activism-based, such as Chocolate Remix.

[^29]: Egbert, et al.

Admittedly, corpus creation also inevitably incurred decisions on the exclusion of material. For instance, I took the decision to not include sapphic artists from Spain, except when in collaboration with Latin artists.[^30] Although several Spanish women singers are now ubiquitous in the Latin music mainstream, the historical development of Spanish women’s music is markedly different from that of Latin America.[^31] Therefore, their inclusion would introduce noise and skew the corpus, which exceeds the scope of this study. Artists not markedly within urban genres, but rather in indie variations, were also not included in the scope of this study. The sampling strategy was also affected by the fact that most lyrics were scraped from Genius, and therefore limited by the songs available on the platform. However, an effort was made to manually retrieve lyrics to lesser-known songs on other online sites, such as AZlyrics. No songs were transcribed by ear, although this remains a viable route in future studies. Lastly, a sample of songs by men was also included in the corpus, to facilitate comparative analysis. This sample was stratified to match the nationalities of the women sub-corpus, since different Latin American countries can bring drastically different vocabulary and slang to song-writing.[^32] Likewise, the sub-corpus includes lesser-known male artists such as Dani Flow.

[^30]: These include Rosalía and Tokischa’s “Linda,” and Nicki Nicole, Snow Tha Product, and Ptazeta’s “Tengo To’.”

[^31]: Simply put, there are a lot of female singers in Spain who sing sapphic lyrics than in Latin America, and their affiliations to the urban Latin sphere are blurry, making their inclusion an issue of representativity.

[^32]: Vankova.

The resulting corpus contains 166 songs, of which 83 are predominantly written by women, and 83 are written by men.[^33] 

[^33]: It is important to mention that not all singers in the corpus were the only songwriters of their songs. Some songs are written in collaboration with producers and other artists. However, the songs selected are ones that are evidently written predominantly by the (woman) performer.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: songs-by-country

corpus_metadata <- 
  metadata %>% 
  mutate(artist_gender = as_factor(ifelse(is_male == 0, "Woman", "Man")))

corpus_metadata %>% 
  ggplot(aes(fct_infreq(artist_country), fill = artist_gender)) +
  geom_bar(position = "dodge") +
  labs(x = "Artist Country", y = "Frequency", fill = "Artist Gender",
       caption = "Fig. 1. Distribution of Songs in the Corpus by Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = c("#561F37", "#39A2AE"), labels = c("Women+", "Men"))

  
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: false
#| label: summary-table

# Summarizing count by artist gender
summary_artist_gender <- corpus_metadata %>% 
  group_by(artist_gender) %>% 
  summarize(count = n())

# Summarizing count and percent by artist country and artist gender
summary_artist_country <- corpus_metadata %>% 
  group_by(artist_country, artist_gender) %>% 
  summarize(count = n(),
            percent = round(count / nrow(corpus_metadata) * 100, 2)) %>% 
  arrange(desc(percent))

# Formatting tables using gt
summary_artist_gender_table <- gt(summary_artist_gender) %>%
  # tab_header(
  #   title = "Summary of Artist Gender",
  #   subtitle = "Count of Artists by Gender"
  # ) %>%
  cols_label(
    artist_gender = "Artist Gender",
    count = "Count"
  ) %>% 
  tab_caption("Fig. 2. Summary of Artists by Gender") %>% 
  tab_options(
    data_row.padding = px(1)
  ) %>% 
  cols_width(everything() ~ px(200)) 

summary_artist_country_table <- gt(summary_artist_country) %>%
  tab_header(
    title = "Summary of Artist Country by Gender",
    subtitle = "Count and Percentage of Artists by Country and Gender"
  ) %>%
  cols_label(
    artist_country = "Artist Country",
    artist_gender = "Artist Gender",
    count = "Count",
    percent = "Percentage"
  ) 

# Printing tables
summary_artist_gender_table
# summary_artist_country_table
  
```

However, this process still raised many questions. For instance, we may be able to determine whether a song is written by a woman, but how can we tell when a song is *about* a woman, and consequently, sapphic? Historically, sapphic singers have utilized different strategies to sing about women without making it apparent. A good example is the case of Ana Gabriel, who since the 1980s has been releasing songs about women, but masked behind male pronouns, such as “Simplemente Amigos” (Just Friends).  Other songs forgo the gendered character of the Spanish language by addressing their lyrics to a “you,” as is the case with Karol G’s “CONTIGO.” In this case, prudent decision-making was implemented. For instance, “CONTIGO” is accompanied by a very explicitly sapphic music video and was therefore included in the corpus. Still, these complications of defining a “sapphic” song merit further discussion, and I encourage future researchers to delve further into this matter.

## 4 **Results**

The prospect of utilizing distant reading tools to understand the "female shift" in urban Latin music has interested previous researchers. In 2022, Angel Torres-Toukoumidis and Isidro Marin-Gutiérrez published a study titled "Computational Analysis of Latin Music Songs Through Tokenization: Case of Female Artists and Reggaetón." This study compiled a corpus of 641 songs by 12 *reggaetoneras* and conducted a lexical analysis of their characteristics. Among their results, Torres-Toukoumidis and Marin-Gutiérrez found that women's reggaetón presented a high index of lexical diversity, where 4992 words were only used five times.[^34] The authors interpreted this finding as counter-evidence to the popular belief that reggaetón is a genre with limited, simple vocabulary. Furthermore, the authors also found a lower usage of zoomorphic characteristics to self-designate, as is common in the genre's discussion of women. Although this study offers an essential introduction to the application of distant reading to urban Latin music, there are many ways in which its methods could be taken a step further.

[^34]: Angel Torres-Toukoumidis, et al. “Computational Analysis of Latin Music Songs Through Tokenization. Case of Female Artists and Reggaeton.” In *Communication and Applied Technologies*, edited by Paulo Carlos López-López, et al., 527–35. Smart Innovation, Systems and Technologies. Singapore: Springer Nature, 2023. <https://doi.org/10.1007/978-981-19-6347-6_47>, p. 531.

This study aims to fill some of these methodological gaps, and will also offer a view into the narrower population of sapphic urban Latin songs. In the first place, this study is based not only on a corpus of songs by women, but also encompasses a comparative sub-corpus of songs by men that will allow to draw direct comparisons between the two groups. Secondly, this study will analyze a particular set of tokens that are commonly used in urban Latin music to denote women and desire, and will compare their usage and discourse across the two groups.

### 4.1. **Frequencies**

An appropriate first step when analyzing token frequency in our corpus, is to visualize a simple word cloud of the most common words (Fig. 2).

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: word-cloud

load("corpus_objects/quanteda_texts_tok.rda")


# create dfm
quanteda_texts_dfm <- dfm(quanteda_texts_tok)


# and we can have a first look at the most frequent words, for instance with a wordcloud

textplot_wordcloud(quanteda_texts_dfm, max_words = 100, color = "black") 
mtext("Fig. 2. 100 Most Common Words in the Corpus",
      side = 1, 
      line = 3, 
      at =-0.07, 
      adj = 0, 
      cex = 0.8)
```

Similar to Fig. 47.1 in Torres-Toukoumidis, et al., the majority of the most common words include common pronouns, articles, and conjunctions found in the Spanish language. In order to make this visualization more helpful to our research goals, we can delete stopwords. Since this analysis uses `quanteda` tools in R, the stopwords deleted were those found in the `stopwords` package. Due to the high usage of English loanwords and expressions in urban Latin music, both Spanish and English words were deleted. Then, the wordclouds were separated by gender.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: word-cloud-by-gender-ns

load("corpus_objects/women_toks.rda")
load("corpus_objects/men_toks.rda")


# create dfm_ns

women_dfm_ns <- dfm(women_toks %>%
                      tokens_remove(pattern = stopwords("es")) %>%
                      tokens_remove(pattern = stopwords("en")))

men_dfm_ns <- dfm(men_toks %>%
                      tokens_remove(pattern = stopwords("es")) %>%
                      tokens_remove(pattern = stopwords("en")))


# and we can have a look at the most frequent words, for instance with a wordcloud

# Generate wordcloud plots
textplot_wordcloud(women_dfm_ns, max_words = 100, color = "#561F37") 
mtext("Fig. 3. 100 Most Common Words in the Women+ Sub-Corpus",
      side = 1, 
      line = 3, 
      at = -0.07, 
      adj = 0, 
      cex = 0.8)

textplot_wordcloud(men_dfm_ns, max_words = 100, color = "#39A2AE")  
mtext("Fig. 4. 100 Most Common Words in the Men Sub-Corpus",
      side = 1, 
      line = 3, 
      at = -0.07, 
      adj = 0, 
      cex = 0.8)
```

In **Fig. 3 and 4**, we can begin to glean some of the particularities of working with urban Latin music specifically. This includes a high number of onomatopeic vocabulary, such as "oh," "wuh," and "trr." Furthermore, for both groups we can see words commonly used to refer to women, such as "baby," and "mami." In both, but perhaps to a slightly more noticeable extent for women, words with corporeal meaning appear, such as "culo," "dura," and "chapa" (ass, toned, ass).

Even though the cleaning of stopwords allows us to look into some of the more specific vocabulary in the corpus, we can clearly see one of the limitations of distant reading tools in working with urban music. Many of the most frequent words in **Fig. 3** **and 4**, are in fact contractions of stopwords that are commonly used in oral settings. These include "na," "pa," and "to'a" which stand for "nada," "para," and "all (feminine)." Since stopword lists currently available for NLP (Natural Language Processing) are predominantly compiled using written language, the tool was unable to detect these contractions.

Furthermore, in much of urban music, it is common for artists to shout out their own names or names of collaborators/producers within their lyrics. In the cleaning process, I made a conscious decision not to take out these so-called "ad-libs," since according to previous literature, a lot of women's voices in urban music have historically been relegated to backing vocals.[^35] However, these instances are visible in the above word clouds, as is the case with "Miko" (for Young Miko) and "Toki" (for Tokischa).

[^35]: Dediós, p. 92.

To experiment with this limitation, I manually compiled a list of tokens to remove from the analyses. These include common contractions of stopwords in Spanish, and names of artists and producers. The list also includes common onomatopeic tokens. A full list of removed words can be found in the `useless_words.txt` file in the Github Repository.

**Fig. 5.** below shows a comparison of the most used tokens for both groups, before and after cleaning. In these plots, the y-axis was converted from token frequency, to a measure of proportion. This is due to the fact that, even though both sub-corpuses contain the same number of songs, the men sub-corpus contained considerably more tokens. This raises interesting questions regarding possible differences in song length or singing speed between the two groups which merits further scholarly inquiry. To avoid the artificial conclusions, proportions were calculated by means of $\frac{Token Frequency}{Total Subgroup Tokens}$

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: prop-graph

frequency_table <- textstat_frequency(quanteda_texts_dfm, groups = is_male)

total_token <- frequency_table %>% 
  group_by(group) %>% 
  count() %>% 
  ungroup() %>% 
  select(n)


frequency_table <- frequency_table %>%
  mutate(token_prop = case_when(
    group == 0 ~ frequency / total_token$n[1],
    group == 1 ~ frequency / total_token$n[2]
  ))

# define useless words
useless_words <- readtext("useless_words.txt", encoding = "UTF-8") %>% 
  tidytext::unnest_lines(input = "text", output = "words")

# proportion table !!
token_prop_table_nw <- frequency_table %>%
  filter(!feature %in% useless_words$words) %>% 
  filter(!feature %in% stopwords(language = "es")) %>% 
  filter(!feature %in% stopwords(language = "en")) %>% 
  group_by(group) %>% 
  arrange(desc(token_prop)) %>% 
  slice_head(n = 20) %>% 
  ungroup() %>% 
  ggplot(aes(x = reorder(feature, token_prop), y = token_prop, color = group)) +
  geom_point() +
  coord_flip() +
  labs(x = NULL, y = "Token Usage Proportion", 
       color = "Artist Gender", title = "Cleaned") +
  scale_color_manual(values = c("#561F37", "#39A2AE"), 
                     labels = c("Women+", "Men")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


token_prop_table_ns <- frequency_table %>%
  filter(!feature %in% stopwords(language = "es")) %>% 
  filter(!feature %in% stopwords(language = "en")) %>% 
  group_by(group) %>% 
  arrange(desc(token_prop)) %>% 
  slice_head(n = 20) %>% 
  ungroup() %>% 
  ggplot(aes(x = reorder(feature, token_prop), y = token_prop, color = group)) +
  geom_point(show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Token Usage Proportion", 
       color = "Artist Gender", title = "Not Cleaned") +
  scale_color_manual(values = c("#561F37", "#39A2AE"), 
                     labels = c("Women+", "Men")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

token_prop_table_ns + token_prop_table_nw +
  labs(caption = "Fig. 5. Most Common Tokens by Gender")
```

In the uncleaned facet, we can observe that some of the onomatopoeic vocabulary seems to occupy a higher proportion of men's vocabulary in the corpus, like in the face of "oh" and "ah." However, other onomatopoeic tokens, such as "prr," are exclusively used by women. So-called "pet names" for women such as "baby" and "mami" seem to occupy a higher proportion of sapphic women's vocabulary. Words exclusively used by men include "party," "baila" (dance), and "cabrón" (Puerto Rican slang to denote 'asshole', gendered male). This could suggest that men are more keen to situate their songs within party/club spaces. Words exclusively used by sapphic women include "bien" (good), "dime" (tell me), and "gusta" (he/she likes). These direct addresses of the object of the song, and a reference to the object's desires, could suggest that songs by women present a higher engagement with the thoughts of the women to which their songs are addressed. For instance, by quite literally asking them to speak.

Whether it is possible to compile a list of words to remove in a way that is objective and does not introduce bias into the dataset is worthy of reflection. The following analyses will work with the entire corpus, including stopwords, since some of the words of interest include female pronouns and articles, such as "ella" (she/her). For a full list of words investigated, see `words_of_interest.txt` .

**Table 1** below shows a number of these results were interesting comparisons were found across both groups. For instance, only the sapphic women subcorpus uses the word "bestie." Perhaps this could be attributed to common storylines of friendship that exist in sapphic songs. In contradiction, it was found that the men subcorpus uses the word "amiga" (female friend) at a higher proportion. This contradiction, and the meaning of these words in their specific context will be investigated further. Relatively similar proportions were found for tokens such as "ella" (her) and "la" (female article). Major differences were found for the words "bitch" and "booty." Along with "bestie," this could suggest a general trend for the usage of words in English to refer to women, especially with regards to the body. As noted in **Fig. 5**, the word "gusta" (he/she likes) remains a notable difference between the two groups.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: token-freq

# look at `words_of_interest`
frequency_table %>% 
  mutate(group = case_when(group == 1 ~ "Men",
                           group == 0 ~ "Women")) %>% 
  rename(Artist_Gender = group) %>% 
  filter(feature %in% c("ella", "la", "bitch", "baby", "gusta", 
                        "conmigo", "booty", "amiga", "bestie")) %>% 
  rename(Token = feature,
         Frequency = frequency,
         Rank = rank,
         Song_Freq = docfreq,
         Proportion = token_prop) %>% 
  mutate(Proportion = round(Proportion, 4)) %>% 
  arrange(Token, Artist_Gender) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 1. Words of Interest Across both Groups"
    )
```

It is important to note, however, that at times these differences can be attributed to specific authors. For instance, the token "bitch" and its overrepresentation in the women's subcorpus can likely be attributed to the artist Young Miko, as evidenced by **Fig. 6**. below.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: by-author


quanteda_texts_dfm %>% 
  textstat_frequency(groups = artist_name) %>% 
  filter(feature == "bitch") %>%
  ggplot(aes(x = group, y = frequency, label = feature)) +
  geom_col(width = .1) +
  # geom_point(size = 1) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7)
  ) +
  labs(y = "Frequency", x = "Artist Name", caption = "Fig. 6. Distribution of the Token 'Bitch' by Artist")

```

### 4.2. Concordances

Another tool in distant reading that can provide further insight into our findings are concordances, which allow us to see the context in which a certain token appears in the corpus.

In **Table 2 and 3** below, I decided to look into one of the most sexualized terms to describe women in urban Latin music– "culo" (ass). As evidenced by **Fig. 3 and 4**, this token was also among the most common in both sub-corpuses. In these contexts, we can see a certain level of lexical imitation from sapphic women. In many of these cases, the word "culo" is used in conjunction with reflexive pronouns such as "me," which emphasize the effect of an action on the speaker. In other words, both sub-corpuses seem to refer to women's bodies in relation to the speaker's wants and desires. Furthermore, in the song "Deseándote" by Nath, the token is placed in cross proximity to "Mercedes," engaging in a discourse of comparison of women with objects.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: concordance-culo

men_toks %>%
  kwic(pattern =  "culo*", window = 5) %>% 
  sample_n(5) %>% 
  as_tibble() %>% 
  select(-from, -to) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 2. Concordance of the Token 'Culo' in Men Sub-Corpus"
    )

women_toks %>%
  kwic(pattern =  "culo*", window = 5) %>% 
  slice(c(1, 5, 18, 31, 40)) %>% # talk about Nath mercedes
  as_tibble() %>% 
  select(-from, -to) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 3. Concordance of the Token 'Culo' in Women Sub-Corpus"
    )
```

Concordances can also help us understand the role of the word "amiga" (female friend) in both groups. In the previous section, we observed how the sapphic subcorpus exclusives uses the word "bestie", but includes a lower proportion of the word "amiga." **Table 4 and 5** reveal a key distinction. In the men subcorpus, the word "amiga" is used to refer to a woman's friends, towards whom the singer is hoping to make sexual advances. In songs such as Bad Bunny's "Titi Me Preguntó," the figure of the female friend is mentioned as a potential "guide" who could advise the woman not to get together with Bad Bunny, because of his commitment issues. By contrast, in several instances of the sapphic subcorpus, the word is used to describe an unsatisfactory friendship that the singer desires to be romantic. Such is the case in Gudnana's AMG "que ladilla aparentar que solo somos amigas" (how upsetting to pretend that we are just friends).

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: concordance-amiga
men_toks %>%
  kwic(pattern =  "amiga*", window = 5) %>% 
  sample_n(5) %>% 
  as_tibble() %>% 
  select(-from, -to) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 4. Concordance of the Token 'Amiga' in Men Sub-Corpus"
    )

women_toks %>%
  kwic(pattern =  "amiga*", window = 5) %>% 
  slice(c(2, 6, 7, 11, 12)) %>% 
  as_tibble() %>% 
  select(-from, -to) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 5. Concordance of the Token 'Amiga' in Women Sub-Corpus"
    )
```

It is also possible to investigate the context of multi-word phrases using concordances. In this case, I was interested in delving further into the differences in the use of reflexivity. For instance, the varying proportion of the token "gusta" (he/she likes) raises questions on the focus on subject versus object in each of these cases. For this experiment, I analyzed the phrases "ella me" and "ella se." The former uses the first-person reflexive pronoun to indicate that the "she" is conducting an action aimed towards the speaker. The latter uses the third-person reflexive pronoun to indicate that she is conducting an action aimed at herself. The first observation we can make is that, while the sapphic subcorpus presents a comparable number of instances of both "ella me" and "ella se," the men's subcorpus presents a notably higher number of "ella me" instances. This could perhaps gesture towards a tendency for more protagonism of the singer in the discourse of the song for the case of the male subcorpus.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: concordance-phrases

# ella me
men_toks %>%
  kwic(pattern = phrase("ella me")) %>% 
  as_tibble() %>% 
  select(-from, -to) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 5. Concordance of the Phrase 'Ella Me' in Men Sub-Corpus"
    )

women_toks %>%
  kwic(pattern = phrase("ella me")) %>% 
  as_tibble() %>% 
  select(-from, -to) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 6. Concordance of the Token 'Ella Me' in Women Sub-Corpus"
    )

# ella se
men_toks %>%
  kwic(pattern = phrase("ella se")) %>% 
  as_tibble() %>% 
  select(-from, -to) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 7. Concordance of the Token 'Ella Se' in Men Sub-Corpus"
    )

women_toks %>%
  kwic(pattern = phrase("ella se")) %>% 
  as_tibble() %>% 
  select(-from, -to) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 8. Concordance of the Token 'Ella Se' in Women Sub-Corpus"
    )
```

### 4.3 **Keyness**

One more test that we can run to look into the matter of lexical comparisons is a keyness test. This tool conducts a Chi-Squared statistical test to determine which words may distinguish a given group from another.

In **Fig. 7**., after removing our list of selected words, we can see some of the words we have already been investigating, such as "gustan," "bestie," and "bitch." However, due to the nature of the corpus, we can also see names that are quite distinctly associated with specific songs, such as "Chulo" and "Brinca." The use of English words such as "booty" and "lick" is noteworthy.

A keyness test is most commonly used to understand distinctions for specific authors relative to a corpus. For the purposes of experimentation, I separated out the Mexican sapphic artist Chzter. In **Fig. 8,** we can see that the key tokens for distinction are often thematic to her specific songs. "Marciana" (Martian) is often used in the song "Mátame Marciana," and the word "lick" is endemic to the song "LICK MY PSSY." Therefore, it is valuable to question how much of these words' distinctiveness can be attributed to individual artists' choices in their personal projects, rather than their adherence to a general movement.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: keyness

# Create a dfm grouped by gender

gender_dfm <- tokens(quanteda_texts_tok, remove_punct = TRUE) %>%
  tokens_remove(stopwords("en")) %>%
  tokens_remove(stopwords("es")) %>%
  tokens_remove(useless_words$words) %>%
  tokens_group(groups = is_male) %>%
  dfm()

# Calculate keyness and determine gender as target group
result_keyness <- textstat_keyness(gender_dfm)

# Plot estimated word keyness
textplot_keyness(result_keyness, color = c("#561F37", "#39A2AE")) +
  labs(caption = "Fig. 7. Keyness Test Across Gender") 



# for instance, we can try with Miko vs the rest

tstat_key <- quanteda_texts_dfm %>% 
  textstat_keyness(target = grepl("Chzter", quanteda_texts_dfm@docvars$docname_))

textplot_keyness(tstat_key, color = c("#561F37", "#39A2AE")) + 
  labs(caption = "Fig. 8. Keyness Test for Chzter")

```

### 4.4 **Lexical Diversity**

For the purpose of investigating the findings of Torres-Toukoumidis, et al., this study also computed measures of lexical diversity across the two groups. This test computes the TTR measure, understood as simply the ratio of token types to the number of tokens in a given text. For this corpus, sapphic women's songs presented a higher TTR as compared to men, which is an interesting addition to the previous findings.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: lexical-diversity

tstat_lexdiv <- quanteda_texts_dfm %>%
  textstat_lexdiv() %>% 
  rename(doc_id = document) %>% 
  left_join(metadata, by = "doc_id") %>% 
  group_by(is_male) %>% 
  summarize(mean_TTR = mean(TTR)) %>% 
  mutate(is_male = case_when(is_male == 1 ~ "Men",
                           is_male == 0 ~ "Women")) %>% 
  rename(Artist_Gender = is_male)


tstat_lexdiv %>% 
  arrange(Artist_Gender) %>% 
  knitr::kable(
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    caption = "Table 9. Lexical Diversity Measures Across Groups"
    )
```

## 5 **Conclusion**

-   Loop theory back in

-   It is hard to make generalizations from the data at hand. The repetition of certain phrases especially if a phrase is part of a chorus, such as in Young Miko's "Wiggy" in **Table 8,** makes it so frequencies can easily become inflated. Token frequencies are helpful in hypothesizing ways in which women could be modifying conventions of the genre, but it is difficult to assess the actuality of these differences. Perhaps a more valuable piece of this analysis is the possibility of understanding how certain commonly-used words take on different meanings in context, such as "amiga." However, since so much of lyrical creativity is a personal choice, it is easier to analyze these characteristics through individual case studies, such as that of "Riri" at the start of this essay. Once more, the domain of sapphic urban Latin music resists canonization, and while we may be able to develop hypotheses, it is hard to find generalizable traits.

-   So much of it is authorial choice.

-   Perhaps we can take this an analyses as a motivation to look further into how the particular experiences of women affect their use of language. With the difficulties in generalization in mind, close-reading analyses can approach these texts with less of an intention to generalize.

## 6 Bibliography

Acosta-Aguilar, Karina. “How Reggaeton Is Enabling Self-Expression, Feminism, and Dismantling Machismo.” *Medium* (blog), May 4, 2023. [https://medium.com/\@karinaacost7/how-reggaeton-is-enabling-self-expression-feminism-and-dismantling-machismo-6462b1a7e588](https://medium.com/@karinaacost7/how-reggaeton-is-enabling-self-expression-feminism-and-dismantling-machismo-6462b1a7e588).

Bode, Katherine. “The Equivalence of ‘Close’ and ‘Distant’ Reading; or, Toward a New Object for Data-Rich Literary History.” *Modern Language Quarterly (Seattle)* 78, no. 1 (2017): 77–106. <https://doi.org/10.1215/00267929-3699787>.

Byszuk, Joanna. “The Voices of Doctor Who – How Stylometry Can Be Useful in Revealing New Information About TV Series.” *Digital Humanities Quarterly* 014, no. 4 (December 20, 2020).

Dediós, Patricio Josué Velarde. “Reguetón de mujeres para mujeres: dos caminos de empoderamiento.” *Antec: Revista Peruana de Investigación Musical* 5, no. 1 (August 14, 2021): 91–109. <https://doi.org/10.62230/antec.v5i1.112>.

Eder, Maciej, Jan Rybicki, and Mike Kestemont. “Stylometry with R: A Package for Computational Text Analysis.” *The R Journal* 8 (January 1, 2016): 107–21. <https://doi.org/10.32614/RJ-2016-007>.

Egbert, Jesse, Douglas Biber, and Bethany Gray. *Designing and Evaluating Language Corpora: A Practical Framework for Corpus Representativeness*. Cambridge University Press, 2022. <https://doi.org/10.1017/9781316584880>.

GitHub. “Distant_Reading_in_R/Text_Analysis/05a_Stylometry.R at Main · ABC-DH/Distant_Reading_in_R.” Accessed March 8, 2024. <https://github.com/ABC-DH/Distant_Reading_in_R/blob/main/Text_Analysis/05a_Stylometry.R>.

GitHub. “Hikmah-Academic-Quarto/Examples/Hikmah-Testing-Custom-Manuscript.Pdf at Main · Andrewheiss/Hikmah-Academic-Quarto.” Accessed March 8, 2024. <https://github.com/andrewheiss/hikmah-academic-quarto/blob/main/examples/hikmah-testing-custom-manuscript.pdf>.

Goldman, Dara E. “Walk like a Woman, Talk like a Man: Ivy Queen’s Troubling of Gender.” *Latino Studies* 15, no. 4 (November 1, 2017): 439–57. <https://doi.org/10.1057/s41276-017-0088-5>.

Grosz, Elizabeth. “Refiguring Lesbian Desire.” In *Space, Time and Perversion*. Routledge, 1995.

Hatzel, Hans Ole, Haimo Stiemer, Chris Biemann, and Evelyn Gius. “Machine Learning in Computational Literary Studies.” *It - Information Technology* 65, no. 4–5 (August 1, 2023): 200–217. <https://doi.org/10.1515/itit-2023-0041>.

Houston, Natalie M. “Distant Reading and Victorian Women’s Poetry.” In *The Cambridge Companion to Victorian Women’s Poetry*, 249–65. Cambridge University Press, 2019. <https://doi.org/10.1017/9781316856543.017>.

LeBrón, Marisol. “‘Con Un Flow Natural’: Sonic Affinities and Reggaeton Nationalism.” *Women & Performance: A Journal of Feminist Theory* 21, no. 2 (July 1, 2011): 219–33. <https://doi.org/10.1080/0740770X.2011.607598>.

Monzingo, Elizabeth, and Daniel Shanahan. “The Expression of Self and Grief in the Nineteenth Century: An Analysis through Distant Readings.” *Nineteenth-Century Music Review* 18, no. 1 (2021): 83–107. <https://doi.org/10.1017/S1479409819000697>.

Mulvey, Laura. “Visual Pleasure and Narrative Cinema.” In *Feminism and Film Theory*, 57–68, 2013. <https://doi.org/10.4324/9780203699362>.

Noriega, Dulce Asela Martínez. “Música, imagen y sexualidad: el reggaetón y las asimetrías de género.” *El Cotidiano*, no. 186 (2014): 63–67.

Rivera-Servera, Ramón H. *Performing Queer Latinidad: Dance, Sexuality, Politics*. 1st ed. Triangulations: Lesbian/Gay/Queer Theater/Drama/Performance. Ann Arbor: University of Michigan Press, 2012. <https://doi.org/10.3998/mpub.2395967>.

Showalter, Elaine. “Feminist Criticism in the Wilderness.” *Critical Inquiry* 8, no. 2 (1981): 179–205.

Solá-Santiago, Frances. “Welcome to Young Miko’s World.” *Rolling Stone* (blog), June 26, 2023. <https://www.rollingstone.com/music/music-features/young-miko-trap-kitty-bad-bunny-karol-g-1234771397/>.

Toro, Ximena de. “Métele con candela pa’ que todas las gatas se muevan. Identidades de género, cuerpo y sexualidad en el reggaetón.” *Revista Punto Género*, no. 1 (October 13, 2011). <https://doi.org/10.5354/2735-7473.2011.16824>.

Torres-Toukoumidis, Angel, Cristian Orlando Coronel Quezada, Jenny Pontón, and Isidro Marín-Gutiérrez. “Computational Analysis of Latin Music Songs Through Tokenization. Case of Female Artists and Reggaeton.” In *Communication and Applied Technologies*, edited by Paulo Carlos López-López, Daniel Barredo, Ángel Torres-Toukoumidis, Andrea De-Santis, and Óscar Avilés, 527–35. Smart Innovation, Systems and Technologies. Singapore: Springer Nature, 2023. <https://doi.org/10.1007/978-981-19-6347-6_47>.

Vankova, Pavlina. “Studying the Vocabulary of Reggaeton Song Lyrics.” *Topics in Linguistics* 23, no. 2 (2022): 63–88. <https://doi.org/10.2478/topling-2022-0012>.
